High geomagnetic field intensity recorded by anorthosite xenoliths requires a vigorous late Mesoproterozoic geodynamo

### 2021-11-08 PNAS_manuscript/PNAS_manuscript.tex
update manuscript

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target \textit{ca.} 1092 Ma anorthosite xenoliths that cooled and acquired their magnetization within the Beaver River diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which has non-ideal paleointensity behaviors leading to its failure in providing reliable paleointensity estimates, the anorthosite xenoliths often yield high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment estimates from the anorthosites is higher than that of the dipole component of Earth's magnetic field today. Such a strong dipole strength recorded by the anorthosites require an existence of a vigorous late Mesoproterozoic geodynamo. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results suggest that a dynamo with strong power sources likely persisted through $\sim$25 Myr at \textit{ca.} 1.1 Ga. These observations are inconsistent with there being a progressive decay of Earth's dynamo strength through the Proterozoic Eon.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB), and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago in the Archean \cite{Selkin2007a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates, which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimates of these values are consistent with an inner core age $>$3 Ga \cite{Gubbins2004a, Konopkova2016a}, other estimates based on first-principles calculations implied higher thermal conductivity values and an age for the inner core that is less than 1.3 Ga \cite{Pozzo2012a, Koker2012a, Gomi2013a, Labrosse2015a, Ohta2016a, Zhang2020b}. With the estimates for the core thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of the core and mantle.

@@

Paleomagnetic data from ancient rocks are one of the few types of observational data that have the potential to record the evolution of Earth’s core and can be used to test the predictions from numerical simulations. Some Precambrian absolute paleointensity records from the PINT database (\url{http://earth.liv.ac.uk/pint/}) show evidence that an appreciable geomagnetic field existed prior to the Ediacaran (Fig. \ref{fig:PINT_compilation}). On the other hand, studies have found very low virtual dipole moment (VDM) estimates from rocks formed in the late Ediacaran \cite{Bono2019a, Shcherbakova2019a, Thallner2021b}, followed by a recovery near the Ediacaran-Cambrian boundary \cite{Thallner2021a}. Other studies focusing on paleomagnetic directions have also found that the late Ediacaran might have been a period of time when the geomagnetic field were experiencing frequent reversals and waning of dipole dominance \cite{Bono2015a, Kodama2020a}. It has been interpreted that these findings are consistent with the results of geodynamo models using high core thermal conductivity values that suggest the inner core nucleation age being near the Ediacaran-Cambrian boundary \cite{Driscoll2016a, Davies2021a}. \cite{Bono2019a} adopted a billion-year-scale view of the Precambrian paleointensity records and based on a polynomial fit through selected data to suggest that the geodynamo power progressively decreased through the Proterozoic (Fig. \ref{fig:PINT_compilation}). That study also recovered very low paleointensity values from \textit{ca.} 565 Ma single silicate crystals and interpreted them to be reflecting the geodynamo to have reached its lowest near the Ediacaran-Cambrian boundary. New low paleointensity estimates derived from the \textit{ca.} 720 Ma Franklin Large Igneous Province are interpreted to support the existence of such decay trend in the Neoproterozoic \cite{Lloyd2021a}.

However, previous paleointensity data from intrusive and extrusive rocks in the Mesoproterozoic Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record Earth's virtual dipole moment (VDM) values similar to that of today, which are higher than the prediction of the long term decay curve (Fig. \ref{fig:PINT_compilation}). \cite{Pesonen1983b} obtained an average virtual dipole moment of 93 $\times 10^21$ Am$^2$ (ZAm$^2$) after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks, including the \textit{ca.} 1108-1105 Ma Osler Volcanics, the \textit{ca.} 1100 Ma Maimainse Point volcanics, \textit{ca.} 1107 Ma intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flow of the \textit{ca.} 1096 Ma North Shore Volcanic Group (Fig. \ref{fig:Geologic_map}). \cite{Kulakov2013a} obtained an average VADM of 59 $\pm$ 11 ZAm$^2$ using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism \cite{Vervoort2007a, Miller2013a}. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by \cite{Pesonen1983b, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. \cite{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas \cite{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by \cite{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VADM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}. This high VADM value estimate from the Midcontinent Rift volcanics may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

Although it is typical of paleointensity experiments to suffer from low success rates, bulk plagioclase-rich rocks \cite{Selkin2000a} and single plagioclase crystals \cite{Tarduno2005a} are found to be able to produce high quality paleointensity results. It has been suggested that plagioclase crystals can protect magnetic minerals that exsolved within them against post-formation weathering and thermochemical alteration during paleointensity experiments \cite{Tarduno2005a}. In this study, we present new paleointensity results from ``IZZI" modified absolute paleointensity experiments \cite{Yu2004a} on a suite of bulk samples from \textit{ca.} 1092 Ma anorthosite xenoliths and their Beaver River diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. Paleointensity experiments on the Beaver River anorthosite xenoliths have a very high success rate. The consistent specimen- and site-level paleointensity results reveal that the magnetite-dominated remanence of the anorthosite specimens has low thermal remanent magnetization (TRM) anisotropy and linear TRM acquisition. Magnetic imaging shows that the anorthosite specimens to have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. With more single-slope, high quality paleointensity data, we seek to compare paleointensity records from the Midcontinent Rift intrusives with those from the extrusives; to test the hypothesis that the Earth's geomagnetic dipole strength remained at high values throughout the Midcontinent Rift; and to further enrich the reliable paleointensity records during the $\sim$25 Myr of Midcontinent Rift magmatic activity.

@@

The Mesoproterozoic North American Midcontinent Rift (MCR) is an intracontinental rift that did not result in the break-up of Laurentia. Although the protracted magmatic activity in the rift lasted form \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}, its activity during the main magmatic stage ($\sim$1098–1090 Ma; \cite{Vervoort2007a, Miller2013a}) was punctuated by rapid and voluminous emplacement of intrusive rocks as well as extrusive volcanics. Based on high-precision $^{206}$Pb/$^{238}$U geochronology and paleomagnetic data, \cite{Swanson-Hysell2021a} hypothesized that the massive Duluth Complex and the coeval North Shore Volcanic group (Fig. \ref{fig:Geologic_map}) formed as a result of upwelling mantled-derived plume material encountering Laurentia's thinned lithosphere, following an ``upside-down” drainage" which rapidly funneled melt to today's northeastern Minnesota (Fig. \ref{fig:Geologic_map}) \cite{Sleep1997a, Swanson-Hysell2014a}). The plagioclase-phyric Anorthositic Series rocks, the crystal-poor Layered Series intrusions of the Duluth Complex, and the coeval volcanics of the North Shore Volcanic Group (NSVG) that outcrop along Northeastern Minnesota (Fig. \ref{fig:Geologic_map}) are interpreted to be a result of such vigorous magmatic activity.

@@

The emplacement of the \textit{ca.} 1092 Ma Beaver Bay Complex punctuates another period of magmatism during the Midcontinent Rift main stage activity. $^{206}$Pb/$^{238}$U zircon dates and cross-cutting relationships tightly constrain the areally extensive Beaver River diabase of the Beaver Bay Complex to have been emplaced at 1091.7 $\pm$ 0.2 Ma \cite{Zhang2021b}. The Beaver River diabase intruded into hypabyssal depths via wide conduits that accommodated the entrainment of numerous anorthosite xenoliths that have short-axis diameters up to 180 meters \cite{Boerboom2004a, Boerboom2006b}. Geochronological, paleomagnetic, geochemical, and petrographic data are consistent with the hypothesis that the Beaver River diabase magma extruded to the surface, feeding the massive Greenstone Flow - one of the largest lava flows on Earth \cite{Doyle2016a, Zhang2021b}.

% The anorthosite xenoliths that distinctly occur within the Beaver River diabase are not only recorders of the wide diabase conduit sizes,  of the geomagnetic field $\sim$1091.7 million years ago. Thermal modeling result and paleomagnetic directional data show the anorthosite xenoliths acquired thermal remanent magnetizations during cooling of the diabase \cite{Zhang2021b}. Step-wise thermal demagnetizations show many anorthosite xenoliths have minimal secondary remanence and single-component characteristic remanence magnetizations that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C.

Despite that the anorthosite xenoliths of the Beaver River diabase were entrained as xenoliths and emplaced only $\sim$4 Myr after the formation of the Anorthositic Series rocks of the Duluth Complex when massive plagioclase mushes were generated, $^{206}$Pb/$^{238}$U zircon geochronology data and zircon rare earth element diffusion modeling results suggest it is the most probable that the Beaver River diabase anorthosite xenoliths are entrained cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. The anorthosite xenoliths of the \textit{ca.} 1092 Ma Beaver River diabase has distinct mineralogy from that of the Anorthositic Series rocks of the Duluth Complex. The dominantly monomineralic anorthosite xenoliths have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. In addition, petrographic imaging support that the plagioclase from the Beaver River anorthosite formed under a magmatic conditions different from the Duluth Complex anorthositic series magma. While the plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture (Fig. \ref{Petro_QDM}A), the more calcium-rich plagioclase of the Beaver River anorthosite often develop granoblastic texture characterized by equigranular plagioclase crystals (Fig. \ref{Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial vermicular titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex Anorthositic Series rocks \cite{Miller1990a}.

@@

Occurrences of magnetite and ilmenite symplectites and needles within anorthositic rocks have been described by \cite{Scofield1975a, Scofield1986a} in the Wichita Mountains and by \cite{Haggerty1976a} in other slowly cooled, mafic intrusions. \cite{Scofield1986a} hypothesized that the inclusions acted as a means of buffering the melt with respect to oxygen, forming either crystallites along preexisting plagioclase substrate or as iron incorporated into the plagioclase that subsequently exsolved at subsolidus temperatures. Such a model requires that plagioclase, instead of other mafic minerals such as olivine, to be the first to crystallize from a magma and act as a sink for iron. The two-stage formation of the Duluth Complex Anorthositic Series fits this model \cite{Miller1990a}. During the first stage, high pressure at subcrustal depths (15-25 km depth) caused fractionation of Duluth Complex magma resulting in the formation of floating plagioclase mushes \cite{Kushiro1980a}. The second stage of crystallization then occurs at low pressures after the magma intruded into shallower chambers.

This proposed magmatic evolution of the Duluth Complex is in distinct contrast to the emplacement of the Beaver River diabase which rapidly ascended and entrained the anorthosite xenoliths directly into shallow chambers \cite{Flower1980a, Miller1997a, Zhang2021b}. Particularly, \cite{Miller1997a} observed a high state of structural disorder within plagioclase of the Beaver River anorthosite using X-ray diffraction analyses. This is consistent with the interpretation that these anorthosite were rapidly transported by upward currents into hypabyssal magma chambers without reaching structural equilibrium. Within the Beaver River anorthosite xenoliths, the titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent (Fig. \ref{Petro_QDM}). Instead, relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Fig. \ref{Petro_QDM}).

@@

40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity selection criteria, yielding 7 anorthosite sites and no diabase sites with site-level paleointensity results that pass paleointensity selections. Example NRM/TRM plots 
\cite{Arai1963a} are shown in Fig. \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are shown in Table \ref{tab:PINT_result} and are plotted in Fig. \ref{fig:PINT_cooling_corrected}. All measurement-level paleointensity experiment data can be accessed through the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED).

@@

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{Paleointensity_plot_cooling_corrected.pdf}
\centering
\caption{\small{Summary plot of cooling rate corrected individual specimen absolute paleointensity results (square and diamond symbols) from this study and their averages at site (grey bar) and locality level (weighted mean by site-level standard deviations; orange bar). }}
\label{fig:PINT_cooling_corrected}
\end{figure*}

>>>>

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target \textit{ca.} 1092 Ma anorthosite xenoliths that cooled and acquired their magnetization within the Beaver River diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which has non-ideal paleointensity behaviors leading to its failure in providing reliable paleointensity estimates, the anorthosite xenoliths often yield high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites is higher than that of the dipole component of Earth's magnetic field today. Such a strong late Mesoproterozoic dipole recorded by the anorthosites require the existence of a vigorous geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results suggest that a dynamo with strong power sources likely persisted through $\sim$25 Myr at \textit{ca.} 1.1 Ga. These observations are inconsistent with there being a progressive decay of Earth's dynamo strength through the Proterozoic Eon.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB), and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago in the Archean \cite{Selkin2007a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates, which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimates of these values are consistent with an inner core age $>$3 Ga \cite{Gubbins2004a, Konopkova2016a}, other estimates based on first principles calculations implied higher thermal conductivity values and an age for the inner core that is less than 1.3 Ga \cite{Pozzo2012a, Koker2012a, Gomi2013a, Zhang2020b}, with some suggesting even younger ages ($<$ 700 Ma) \cite{Labrosse2015a, Ohta2016a}. With the estimates for the core's thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of the core and mantle.

@@

Paleomagnetic data from ancient rocks are one of the few types of observational data that have the potential to record the evolution of Earth’s core and can be used to test the predictions from numerical simulations. Previous interpretations of the current Precambrian absolute paleointensity records (compiled in the PINT database; \url{http://earth.liv.ac.uk/pint/}) suggest that an appreciable geomagnetic field existed in the Archean and progressively decayed through the Proterozoic till it reached its weakest in the Ediacaran when very low virtual dipole moment (VDM) estimates have been obtained \cite{Bono2019a, Shcherbakova2019a, Thallner2021b, Lloyd2021a}, followed by a recovery near the Ediacaran-Cambrian boundary \cite{Thallner2021a}. Bono et al \cite{Bono2019a} adopted a billion-year-scale view of such observations of the Precambrian geomagnetic field intensities to suggest that these evidence are consistent with models that use high core thermal conductivity values where the nucleation of the inner core began to provide compositional convection power for driving the geodynamo during the Ediacaran-Cambrian after experiencing a progressive decay when the geodynamo was dominantly driven by thermal conduction earlier in the Proterozoic \cite{Driscoll2016a, Davies2021a}.

% Other studies focusing on paleomagnetic directions have also found that the late Ediacaran might have been a period of time when the geomagnetic field were experiencing frequent reversals and waning of dipole dominance \cite{Bono2015a, Kodama2020a}.

However, previous paleointensity data from intrusive and extrusive rocks in the \textit{ca.} 1.1 Ga Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record VDM values similar to that of today, which are higher than the values predicted by the long term decay curve proposed by Bono et al \cite{Bono2019a} (Fig. \ref{fig:PINT_compilation}). Pesonen and Halls \cite{Pesonen1983b} obtained an average virtual dipole moment of 93 $\times 10^21$ Am$^2$ (ZAm$^2$) after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks from the Midcontinent Rift, including the \textit{ca.} 1108-1105 Ma Osler Volcanics, the \textit{ca.} 1100 Ma Maimainse Point volcanics, \textit{ca.} 1107 Ma intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flows of the \textit{ca.} 1096 Ma North Shore Volcanic Group (Fig. \ref{fig:Geologic_map}). Kulakov et al \cite{Kulakov2013a} obtained an average VDM of 59 $\pm$ 11 ZAm$^2$ using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism \cite{Vervoort2007a, Miller2013a}. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by ref. \citealp{Pesonen1983b, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. Pesonen and Halls \cite{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas Kulakov et al \cite{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by Sprain et al \cite{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VDM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}. This high value from the Midcontinent Rift volcanics may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

Although it is typical of paleointensity experiments to suffer from low success rates, bulk plagioclase-rich rocks \cite{Selkin2000a} and single plagioclase crystals \cite{Tarduno2005a} are found to be able to produce high quality paleointensity results. It has been suggested that plagioclase crystals can protect magnetic minerals that exsolved within them against post-formation weathering and thermochemical alteration during paleointensity experiments \cite{Tarduno2005a}. In this study, we present new paleointensity results from ``IZZI" modified absolute paleointensity experiments \cite{Yu2004a} on a suite of bulk samples from \textit{ca.} 1092 Ma anorthosite xenoliths and their Beaver River diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. Paleointensity experiments on the anorthosite xenoliths have a very high success rate. Their consistent specimen- and site-level paleointensity results together with rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. With more single-slope, high quality paleointensity data, further enrich the reliable paleointensity records during the $\sim$25 Myr of Midcontinent Rift magmatic activity. Our new data from the Midcontinent Rift intrusive rocks support previous data from the volcanics that there was a strong late Mesoproterozoic geodynamo.

@@

The Mesoproterozoic North American Midcontinent Rift is an intracontinental rift that did not result in the break-up of Laurentia. Although the protracted magmatic activity in the rift lasted form \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}, its activity during the main magmatic stage ($\sim$1098–1090 Ma; \cite{Vervoort2007a, Miller2013a}) was characterized by rapid and voluminous emplacement of intrusive rocks and extrusive volcanics. The $<$500 Kyr emplacement of the \textit{ca.} 1096 Ma Duluth Complex and the $~$8 km thick volcanics associated with it is one manifestation \cite{Swanson-Hysell2021a}. The emplacement of the \textit{ca.} 1092 Ma Beaver Bay Complex punctuates another such period of magmatism during the Midcontinent Rift main stage activity. High precision $^{206}$Pb/$^{238}$U zircon dates from one anorthosite xenolith and units that have cross-cutting relationships with the Beaver River diabase tightly constrain the timing of diabase emplacement to be at 1091.7 $\pm$ 0.2 Ma \cite{Zhang2021b}. The xenolith-bearing diabase dikes and sills intruded into hypabyssal depths via wide conduits that accommodated the entrainment of numerous anorthosite xenoliths that have short-axis diameters ranging from centimeters to up to 180 meters \cite{Boerboom2004a, Boerboom2006b}.

Previous geochronology data and rare earth element diffusion modeling suggest that the anorthosite xenoliths of the Beaver River diabase are cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. Petrological textures and geochemical attributes imply that the Beaver River anorthosite xenoliths formed in a lower crustal magmatic condition distinct from that of the Duluth Complex which is $\sim$4 Myr older and also generated massive anorthositic cumulates that led to the formation of the Anorthositic Series \cite{Zhang2021b, Swanson-Hysell2021a}. The dominantly monomineralic anorthosite xenoliths of the Beaver River diabase have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. The plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture showing igneous foliations (Fig. \ref{Petro_QDM}A), but the more calcium-rich plagioclase of the Beaver River anorthosites often develop granoblastic texture characterized by equigranular crystals with poor alignments (Fig. \ref{Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial symplectic titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex Anorthositic Series rocks \cite{Miller1990a}. Such a high concentration of Fe oxides carrying magnetic remanence in the slowly cooled Duluth Complex anorthositic rocks reflect that iron in the plagioclase mush in the magma chamber acted as a means of buffering the melt with respect to oxygen, forming crystallites along preexisting plagioclase substrate or as iron oxides incorporated into the plagioclase that subsequently exsolved at subsolidus temperatures \cite{Scofield1986a}. In contrast, titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent but relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Fig. \ref{Petro_QDM}). This likely reflects that the Beaver River diabase magma had a different magma oxidation history than that of the Duluth Complex. The plagioclase and oxide mineral textures of the Beaver River anorthosites also distinguish them from plagioclase cumulates of other layered mafic complex where magnetic mineral fabrics associated with igneous foliation often occur \cite{Scofield1986a, Selkin2000a, Feinberg2006a}.

% the Beaver River diabase rapidly ascended and entrained the anorthosite xenoliths directly into shallow chambers and likely did not reach  \cite{Flower1980a, Miller1997a, Zhang2021b}. Particularly, \cite{Miller1997a} observed a high state of structural disorder within plagioclase of the Beaver River anorthosite using X-ray diffraction analyses. This is consistent with the interpretation that these anorthosite were rapidly transported by upward currents into hypabyssal magma chambers without reaching structural equilibrium. Within the Beaver River anorthosite xenoliths, the titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent (Fig. \ref{Petro_QDM}). Instead, relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Fig. \ref{Petro_QDM}).

% In the Beaver River anorthosite xenoliths, relatively sparse titano-magnetite occur within or interstitial to the plagioclase (Fig. \ref{Petro_QDM}).

% Occurrences of magnetite and ilmenite symplectites and needles within anorthositic rocks have been described by \cite{Scofield1975a, Scofield1986a} in the Wichita Mountains and by \cite{Haggerty1976a} in other slowly cooled, mafic intrusions. \cite{Scofield1986a} hypothesized that the inclusions acted as a means of buffering the melt with respect to oxygen, forming either crystallites along preexisting plagioclase substrate or as iron incorporated into the plagioclase that subsequently exsolved at subsolidus temperatures. Such a model requires that plagioclase, instead of other mafic minerals such as olivine, to be the first to crystallize from a magma and act as a sink for iron. The two-stage formation of the Duluth Complex Anorthositic Series fits this model \cite{Miller1990a}. During the first stage, high pressure at subcrustal depths (15-25 km depth) caused fractionation of Duluth Complex magma resulting in the formation of floating plagioclase mushes \cite{Kushiro1980a}. The second stage of crystallization then occurs at low pressures after the magma intruded into shallower chambers.

% This proposed magmatic evolution of the Duluth Complex is in distinct contrast to the emplacement of the Beaver River diabase which rapidly ascended and entrained the anorthosite xenoliths directly into shallow chambers \cite{Flower1980a, Miller1997a, Zhang2021b}. Particularly, \cite{Miller1997a} observed a high state of structural disorder within plagioclase of the Beaver River anorthosite using X-ray diffraction analyses. This is consistent with the interpretation that these anorthosite were rapidly transported by upward currents into hypabyssal magma chambers without reaching structural equilibrium. Within the Beaver River anorthosite xenoliths, the titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent (Fig. \ref{Petro_QDM}). Instead, relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Fig. \ref{Petro_QDM}).

% it is the most probable that the anorthosite xenoliths of the Beaver River diabase are cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. Thermal modeling results and paleomagnetic directional data show the anorthosite xenoliths acquired thermal remanent magnetizations during cooling of the diabase \cite{Zhang2021b}. Step-wise thermal demagnetizations show many anorthosite xenoliths have minimal secondary remanence and single-component characteristic remanence magnetizations that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C.

% Geochronological, paleomagnetic, geochemical, and petrologic data are consistent with the hypothesis that the Beaver River diabase magma extruded to the surface, feeding the massive Greenstone Flow - one of the largest lava flows on Earth \cite{Doyle2016a, Zhang2021b}.

% Based on high-precision $^{206}$Pb/$^{238}$U geochronology and paleomagnetic data, \cite{Swanson-Hysell2021a} hypothesized that the massive Duluth Complex and the coeval North Shore Volcanic group (Fig. \ref{fig:Geologic_map}) formed as a result of upwelling mantled-derived plume encountering Laurentia's thinned lithosphere, following an ``upside-down” drainage" which rapidly funneled melt to today's northeastern Minnesota (Fig. \ref{fig:Geologic_map}) \cite{Sleep1997a, Swanson-Hysell2014a}). The plagioclase-phyric Anorthositic Series rocks, the crystal-poor Layered Series intrusions of the Duluth Complex, and the coeval volcanics of the North Shore Volcanic Group (NSVG) that outcrop along Northeastern Minnesota (Fig. \ref{fig:Geologic_map}) are interpreted to be a result of such vigorous magmatic activity.

@@

% Despite that the anorthosite xenoliths of the Beaver River diabase were entrained as xenoliths and emplaced only $\sim$4 Myr after the formation of the Anorthositic Series rocks of the Duluth Complex when massive plagioclase mushes were generated, $^{206}$Pb/$^{238}$U zircon geochronology data and zircon rare earth element diffusion modeling results suggest it is the most probable that the Beaver River diabase anorthosite xenoliths are entrained cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. The anorthosite xenoliths of the \textit{ca.} 1092 Ma Beaver River diabase has distinct mineralogy from that of the Anorthositic Series rocks of the Duluth Complex. The dominantly monomineralic anorthosite xenoliths have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. In addition, petrographic imaging support that the plagioclase from the Beaver River anorthosite formed under a magmatic conditions different from the Duluth Complex anorthositic series magma. While the plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture (Fig. \ref{Petro_QDM}A), the more calcium-rich plagioclase of the Beaver River anorthosite often develop granoblastic texture characterized by equigranular plagioclase crystals (Fig. \ref{Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial vermicular titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex Anorthositic Series rocks \cite{Miller1990a}.

@@

@@

40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity selection criteria, yielding 7 anorthosite sites and no diabase sites with site-level paleointensity results that pass paleointensity selections. Example NRM/TRM plots 
\cite{Arai1963a} are shown in Fig. \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are shown in Table S1 and are plotted in Fig. \ref{fig:PINT_cooling_corrected}. All measurement-level paleointensity experiment data can be accessed through the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED).

@@

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{Paleointensity_plot_cooling_corrected.pdf}
\centering
\caption{\small{Summary plot of individual specimen absolute paleointensity results (square and diamond symbols) from this study and their averages at site (grey bar) and locality level (weighted mean by site-level standard deviations; orange bar). All results are corrected for cooling rate. }}
\label{fig:PINT_cooling_corrected}
\end{figure*}
